Well-quasi-orderings on word languages

### 2024-08-12 src/infixes.tex
trying to add amalgamation

\begin{definition}
    \textbf{TODO}. An \intro{amalgamation system}
    is a triple such that.
\end{definition}

Let us recall some examples of \kl{amalgamation systems} given by the authors
of the original paper: regular languages \cite[Theorem 5.3]{ASZZ24}, VASS as a
consequence of \cite[Theorem 5.5]{ASZZ24}, context-free languages as a
consequence of \cite[Theorem 5.10]{ASZZ24}.

\begin{lemma}
    Let $L$ be a language recognized by an \kl{amalgamation system}
    and assume that $L$ is \kl{well-quasi-ordered}
    by the \kl{infix relation}. Then,
    for every minimal run $\rho$,
    the \kl{gap languages} of $\rho$
    are \kl{subsemigroup languages}.
\end{lemma}

\begin{proofof}{infix-amalgamation:thm}
    TODO.
\end{proofof}

\begin{lemma}
    Let $L$ be given by an \kl{effective amalgamation system},
    then it is decidable whether $L$ 
    is \kl{well-quasi-ordered} by the \kl{infix relation}.
\end{lemma}

>>>>

\AP Let us now formally introduce the notion of \kl{amalgamation systems}, and
recall some results from \cite{ASZZ24} that will be useful for the proof of
\cref{infix-amalgamation:thm}. Let us recall that over an alphabet $(\Sigma,
=)$ a \kl{subword embedding} between two words $u \in \Sigma^*$ and $v \in
\Sigma^*$ is a function $\rho \colon \range{\card{u}} \to \range{\card{v}}$
such that $u_i = v_{\rho(i)}$ for all $i \in \range{\card{u}}$. We write
$\intro*\HigEmb(u,v)$ the set of all \kl{subword embeddings} between $u$ and
$v$. It may be useful to notice that the set of finite words over $\Sigma$
forms a category when we consider \kl{subword embeddings} as morphisms.

\AP Given a \kl{subword embedding} $f \colon u \to v$ between two words $u$ and
$v$, there exists a unique decomposition $v = \GapWord{f}{0} u_1 \GapWord{f}{1}
\cdots \GapWord{f}{k-1} u_k \GapWord{f}{k}$ where $\GapWord{f}{i} =
v[f(i)+1:f(i+1)-1]$ for all $1 \leq i \leq k-1$, $\GapWord{f}{k} =
v[f(k)+1:\card{v}]$, and $\GapWord{f}{0}   = v[1: f(1)-1]$.  We say that
$\intro*\GapWord{f}{i}$ is the $i$-th \intro{gap word} of $f$. Given a run $r
\in R$, and $i \in \range[0]{\card{\canrun(r)}}$, we define the \intro{gap
language} of $r$ at position $i$ $\GapLanguage{r}{i} \defined
\setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.

\begin{definition}
    An \intro{amalgamation system}
    is a triple $(\Sigma, R, E, \canrun)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    and 
    $E$ is a set of so-called \emph{run embeddings}.
    \begin{itemize}
        \item For all $r, s \in R$,
            $E(r,s) \subseteq \HigEmb(\canrun(r), \canrun(s))$.
        \item $(R, E)$ forms a category.
        \item $(R, \leq_E)$ is a well-quasi-ordered set.
        \item concatenative amalagamation
    \end{itemize}

    The \intro{amalgamation language} of such a system
    is the set of all words $w \in \Sigma^*$ such that
    there exists a run $r \in R$
    such that the concatenation of the letters (and possibly the empty word) in
    $\canrun(r)$ equals $w$.
\end{definition}

The notion of \kl{amalgamation system} is tailored to produce \emph{pumping
arguments}, which is exactly what our \cref{pumping-periods:lem} talks about.
Because the notion is quite abstract, it may be useful to replicate some
intuitions developed by the authors of \cite{ASZZ24}.

Let us recall some examples of languages that can be recognized by
\kl{amalgamation systems} given by the authors of the original paper: regular
languages \cite[Theorem 5.3]{ASZZ24}, VASS as a consequence of \cite[Theorem
5.5]{ASZZ24}, context-free languages as a consequence of \cite[Theorem
5.10]{ASZZ24}.

\begin{proofof}{infix-amalgamation:thm}
    Assume that $L$ is \kl{well-quasi-ordered} by the \kl{infix relation},
    and obtained by an \kl{amalgamation system}
    $(\Sigma, R, E, \canrun)$.

    Let us consider the set $M$ of minimal runs for the relation $\leq_E$,
    which is finite because the latter is a \kl{well-quasi-ordering}. By
    definition, for every word $w \in L$, there exists $r \in M$, $s \in R$,
    and $f \in E(r,s)$ such that $w = \canrun(s)$.
    In particular, we conclude that
    \begin{equation*}
        L \subseteq \bigcup_{r \in M} 
        \left(
        \GapLanguage{r}{0}
        \prod_{i = 1}^{\card{\canrun(r)}-1} \canrun(r)_i \GapLanguage{r}{i} 
        \right)
        \quad .
    \end{equation*}

    Let us first argue that languages $\GapLanguage{r}{i}$ are all \kl{chains}
    for the \kl{prefix ordering}. According to \cite[Lemma 4.1]{ASZZ24}, either
    $\GapLanguage{r}{i}$ is a \kl{chain} for the \kl{prefix relation}, or it is
    an \kl{unbounded language}. Note that \kl{unbounded languages}
    are \kl{antichains} for the \kl{infix relation}, and therefore
    \textbf{unclear}.

    Then, we can leverage \cref{pumping-periods:lem}
    to conclude that $L$ is a finite union of \kl{chains}
    for the \kl{infix relation}.
\end{proofof}

\begin{conjecture}
    Let $L$ be given by an \kl{effective amalgamation system},
    then it is decidable whether $L$ 
    is \kl{well-quasi-ordered} by the \kl{infix relation}.
\end{conjecture}

### 2024-08-12 src/introduction.tex
amalgamation?
trying to add amalgamation

Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$,
written $u \intro*\higleq v$,
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. In our, case we
consider alphabets $\Sigma$ that are finite and equipped with the equality
relation. In this setting, the three alternatives we consider are the
\intro{prefix relation} --- $u \intro*\prefleq v$ if there exists $w$ with $uw
= v$ ---, the \intro{suffix relation} --- $u \intro*\suffleq v$ if there exists
$w$ such that $wu = v$ ---, and the \intro{infix relation} --- $u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$ ---.
Unfortunately, it is an easy check that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters.
This is not the end of the road, because some \emph{subsets} of $\Sigma^*$
might still be \kl{well-quasi-ordered} by these relations, the simplest example
being the case of finite sets of (finite) words which are all
\kl{well-quasi-ordered} regardless of the ordering considered. This raises the
question of characterizing the subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} with respect to the \kl{prefix relation} (respectively,
the \kl{suffix relation} or the \kl{infix relation}), and providing suitable
decision procedures.

\subparagraph{Contribution} In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Let us remark that these word orderings are all
\intro{well-founded}, i.e., they do not contain infinite \intro{decreasing
sequences} $\seqof{x_n}$ such that $x_{n+1} \prec x_n$ for all $n \in \Nat$.
Let us recall that the key difference between \kl{well-quasi-orderings} and the
more usual notion of well-founded orderings is that the latter allows the
existence of infinite \intro{antichains}, that is, sets of pairwise
incomparable elements. In fact, a set $X$ is \kl{well-quasi-ordered} if and
only if it is \kl{well-founded} and does not contain an infinite \kl{antichain}
\cite{KRU72}.  As a consequence, our study focuses on detecting infinite
\kl{antichains} in languages $L \subseteq \Sigma^*$.

>>>>

\AP
Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$,
written $u \intro*\higleq v$,
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

\AP
However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. In our, case we
consider alphabets $\Sigma$ that are finite and equipped with the equality
relation. In this setting, the three alternatives we consider are the
\intro{prefix relation} --- $u \intro*\prefleq v$ if there exists $w$ with $uw
= v$ ---, the \intro{suffix relation} --- $u \intro*\suffleq v$ if there exists
$w$ such that $wu = v$ ---, and the \intro{infix relation} --- $u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$ ---.
Unfortunately, it is an easy check that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters.
This is not the end of the road, because some \emph{subsets} of $\Sigma^*$
might still be \kl{well-quasi-ordered} by these relations, the simplest example
being the case of finite sets of (finite) words which are all
\kl{well-quasi-ordered} regardless of the ordering considered. This raises the
question of characterizing the subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} with respect to the \kl{prefix relation} (respectively,
the \kl{suffix relation} or the \kl{infix relation}), and providing suitable
decision procedures.

\subparagraph{Contribution} 
\AP
In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Let us remark that these word orderings are all
\intro{well-founded}, i.e., they do not contain infinite \intro{decreasing
sequences} $\seqof{x_n}$ such that $x_{n+1} \prec x_n$ for all $n \in \Nat$.
Let us recall that the key difference between \kl{well-quasi-orderings} and the
more usual notion of well-founded orderings is that the latter allows the
existence of infinite \intro{antichains}, that is, sets of pairwise
incomparable elements. In fact, a set $X$ is \kl{well-quasi-ordered} if and
only if it is \kl{well-founded} and does not contain an infinite \kl{antichain}
\cite{KRU72}.  As a consequence, our study focuses on detecting infinite
\kl{antichains} in languages $L \subseteq \Sigma^*$.
